---
title: "DA Project"
author: "Rathnapriya Gopalakrishnan"
date: "4/21/2022"
output:
  word_document: default
  html_document: default
---
# Data Storage 
## Importing Dataset
```{r}
df = read.csv("C:/Users/Priya/Desktop/heart_2020_cleaned.csv")
head(df)
unalteredDf = df
```
# Data Cleaning

## Checking for missing values
```{r}
sum(is.na(df))
```

No missing values to handle

# Data Exploration
```{r}
str(df)
```

```{r}
summary(df)
```

# Feature Engineering
## Extracting the continuous and categorical features

```{r}
numeric_df <- df[, sapply(df, class) == "numeric"]
head(numeric_df)
```

```{r}
cat_df <- df[, sapply(df, class) == "character"]
head(cat_df)
```
```{r}
df$Smoking = as.factor(df$Smoking)
df$AlcoholDrinking = as.factor(df$Smoking)
df$Stroke = as.factor(df$Stroke)
df$DiffWalking = as.factor(df$DiffWalking)
df$Sex = as.factor(df$Sex)
df$AgeCategory = as.factor(df$AgeCategory)
df$Race = as.factor(df$Race)
df$Diabetic = as.factor(df$Diabetic)
df$PhysicalActivity = as.factor(df$PhysicalActivity)
df$GenHealth = as.factor(df$GenHealth)
df$Asthma = as.factor(df$Asthma)
df$KidneyDisease = as.factor(df$KidneyDisease)
df$SkinCancer = as.factor(df$SkinCancer)
#df$HD = ifelse (df$HeartDisease=="Yes",1,0)
df$HeartDisease = factor(df$HeartDisease,levels = c("No","Yes"),labels = c(0,1))
```

```{r}
head(df$HeartDisease)
```



```{r}
str(df)
```



```{r}
cor(numeric_df)
```

```{r}
library(corrplot)
corrplot(cor(numeric_df),method='color')
```


# Data Visualization

## For continuous variables
```{r}
boxplot(numeric_df)
```

## How each continuous variable differs for different values of Heart Disease

```{r}
par(mfrow=c(2,2))
boxplot(df$BMI~ df$HeartDisease)
boxplot(df$PhysicalHealth~ df$HeartDisease)
boxplot(df$MentalHealth~ df$HeartDisease)
boxplot(df$SleepTime~ df$HeartDisease)
```


## Histogram of continuous variables
```{r}
par(mfrow=c(2,2))
hist(numeric_df$BMI)
hist(numeric_df$PhysicalHealth)
hist(numeric_df$MentalHealth)
hist(numeric_df$SleepTime)
```

## For categorical variables

## Frequency of classes for each categorical variable
```{r}
library(ggplot2)
ggplot(df) + geom_bar(aes(x = HeartDisease, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = Smoking, fill=Smoking))
ggplot(df) + geom_bar(aes(x = AlcoholDrinking, fill=AlcoholDrinking))
ggplot(df) + geom_bar(aes(x = Stroke, fill=Stroke))
ggplot(df) + geom_bar(aes(x = DiffWalking, fill=DiffWalking))
ggplot(df) + geom_bar(aes(x = Sex, fill=Sex))
ggplot(df) + geom_bar(aes(x = AgeCategory, fill=AgeCategory))
ggplot(df) + geom_bar(aes(x = Race, fill=Race))
ggplot(df) + geom_bar(aes(x = Diabetic, fill=Diabetic))
ggplot(df) + geom_bar(aes(x = PhysicalActivity, fill=PhysicalActivity))
ggplot(df) + geom_bar(aes(x = GenHealth, fill=GenHealth))
ggplot(df) + geom_bar(aes(x = Asthma, fill=Asthma))
ggplot(df) + geom_bar(aes(x = KidneyDisease, fill=KidneyDisease))
ggplot(df) + geom_bar(aes(x = SkinCancer, fill=SkinCancer))
```


## Frequency of Heart Disease in each class of categorical variables 
```{r}
library(ggplot2)
ggplot(df) + geom_bar(aes(x = HeartDisease, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = Smoking, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = AlcoholDrinking, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = Stroke, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = DiffWalking, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = Sex, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = AgeCategory, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = Race, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = Diabetic, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = PhysicalActivity, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = GenHealth, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = Asthma, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = KidneyDisease, fill=HeartDisease))
ggplot(df) + geom_bar(aes(x = SkinCancer, fill=HeartDisease))
```

# Sampling
## We have an imbalance in the data

No of records in each class of the target variable before sampling
```{r}
table(df$HeartDisease)
```

## Performing Undersampling

```{r}
library(ROSE)
library(TeachingDemos)
set.seed(char2seed("Priya"))
balanced_df_under = ovun.sample(HeartDisease~., 
                          data=df,
                          method = "under")$data
```

No of records in each class of the target variable after undersampling
```{r}
table(balanced_df_under$HeartDisease)
```

## Performing Oversampling
```{r}
library(TeachingDemos)
set.seed(char2seed("Priya"))
balanced_df_over = ovun.sample(HeartDisease~., 
                          data=df,
                          method = "over")$data
```


No of records in each class of the target variable after oversampling
```{r}
table(balanced_df_over$HeartDisease)
```



```{r}

MLdf=balanced_df_under


# MLdf0 = subset(MLdf, HeartDisease==0)
# MLdf1 = subset(MLdf, HeartDisease==1)
# MLdf0 = MLdf0[1:500,]
# MLdf1 = MLdf1[1:500,]
# 
# MLdf = rbind(MLdf0, MLdf1)


# MLdf=balanced_df_under[1:1000,]

nrow(MLdf)
ncol(MLdf)
```


# Machine Learning
## Forming train and test sets
```{r}
library(TeachingDemos)
set.seed(char2seed("Priya"))
train = sample(nrow(MLdf), size = 0.8*nrow(MLdf))
head(MLdf)
```

Removing HD column
```{r}
#MLdf = subset(MLdf, select = -c(HD))
```

```{r}
trainDf = MLdf[train,]
testDf = MLdf[-train,]
```

Extracting Train and Test sets
```{r}
trainMldf = MLdf[train,]
testMldf = MLdf[-train,]
Xtrain = trainDf[,-c(1)]
Xtest =  testDf[,-c(1)]
ytrain = trainDf$HeartDisease
ytest = testDf$HeartDisease
```


## Logistic Regression

```{r}
head(MLdf)
```

```{r}
str(MLdf)
```


```{r}
library(gbm)

log_mod = glm(HeartDisease~., data=trainMldf, family="binomial")
summary(log_mod)

```


Function to extract evaluation metrics
```{r}

met <- function(cm){
  print(cm)
  met_list = list( "Acc" =cm$overall['Accuracy'], "Spec" = cm$byClass['Specificity'], "Sens" = cm$byClass['Sensitivity'])
  return (met_list)
}

```




```{r}
proba = predict(log_mod, testMldf, type = "response")
pred = ifelse(proba>0.5, 1,0)

```

```{r}
mean(pred!=testMldf$HeartDisease)
```


```{r}
library(caret)

pred = factor(pred, levels=c(0,1), labels=c("X0","X1"))
testMldf$HeartDisease = factor(testMldf$HeartDisease, levels=c(0,1), labels=c("X0","X1"))

cm_glm= confusionMatrix(pred,testMldf$HeartDisease, positive='X1' )
eval_glm = met(cm_glm)
print(cm_glm)
```



```{r}
library(pROC)
roc_glm = roc(testMldf$HeartDisease , proba)
plot(roc_glm, main = "ROC Curve for Logistic Regression")
```


## Random Forest




Converting train and test categorical variables to factors
```{r}
levels(trainMldf$HeartDisease) <- make.names(levels(factor(trainMldf$HeartDisease)))

levels(trainMldf$Smoking) <- make.names(levels(factor(trainMldf$Smoking)))

levels(trainMldf$AlcoholDrinking) <- make.names(levels(factor(trainMldf$AlcoholDrinking)))

levels(trainMldf$Stroke) <- make.names(levels(factor(trainMldf$Stroke)))

levels(trainMldf$DiffWalking) <- make.names(levels(factor(trainMldf$DiffWalking)))

levels(trainMldf$Sex) <- make.names(levels(factor(trainMldf$Sex)))

levels(trainMldf$AgeCategory) <- make.names(levels(factor(trainMldf$AgeCategory)))

levels(trainMldf$Race) <- make.names(levels(factor(trainMldf$Race)))

levels(trainMldf$Diabetic) <- make.names(levels(factor(trainMldf$Diabetic)))

levels(trainMldf$PhysicalActivity) <- make.names(levels(factor(trainMldf$PhysicalActivity)))

levels(trainMldf$GenHealth) <- make.names(levels(factor(trainMldf$GenHealth)))

levels(trainMldf$Asthma) <- make.names(levels(factor(trainMldf$Asthma)))

levels(trainMldf$KidneyDisease) <- make.names(levels(factor(trainMldf$KidneyDisease)))

levels(trainMldf$SkinCancer) <- make.names(levels(factor(trainMldf$SkinCancer)))
```

```{r}


levels(testMldf$HeartDisease) <- make.names(levels(factor(testMldf$HeartDisease)))

levels(testMldf$Smoking) <- make.names(levels(factor(testMldf$Smoking)))

levels(testMldf$AlcoholDrinking) <- make.names(levels(factor(testMldf$AlcoholDrinking)))

levels(testMldf$Stroke) <- make.names(levels(factor(testMldf$Stroke)))

levels(testMldf$DiffWalking) <- make.names(levels(factor(testMldf$DiffWalking)))

levels(testMldf$Sex) <- make.names(levels(factor(testMldf$Sex)))

levels(testMldf$AgeCategory) <- make.names(levels(factor(testMldf$AgeCategory)))

levels(testMldf$Race) <- make.names(levels(factor(testMldf$Race)))

levels(testMldf$Diabetic) <- make.names(levels(factor(testMldf$Diabetic)))

levels(testMldf$PhysicalActivity) <- make.names(levels(factor(testMldf$PhysicalActivity)))

levels(testMldf$GenHealth) <- make.names(levels(factor(testMldf$GenHealth)))

levels(testMldf$Asthma) <- make.names(levels(factor(testMldf$Asthma)))

levels(testMldf$KidneyDisease) <- make.names(levels(factor(testMldf$KidneyDisease)))

levels(testMldf$SkinCancer) <- make.names(levels(factor(testMldf$SkinCancer)))

```







```{r}
library(randomForest)
library(caret)

library(TeachingDemos)
set.seed(char2seed("Priya"))

x = trainControl(method = "cv", number = 3, classProbs = TRUE, summaryFunction = twoClassSummary)

rf_mod_1 <- train(HeartDisease~. , data = trainMldf, method = "rf", trControl = x,  metric = "ROC")
print(rf_mod_1)
```

```{r}
plot(rf_mod_1)
```


```{r}
mtry = rf_mod_1$bestTune[1,]
rf_mod <- randomForest(HeartDisease ~ . , data= trainMldf, mtry=mtry, importance=TRUE)
summary(rf_mod)

```


```{r}
varImp(rf_mod)
```


```{r}
varImpPlot(rf_mod)
```


```{r}
# testMldf$HeartDisease = factor(testMldf$HeartDisease, levels = levels(trainMldf$HeartDisease))
levels(testMldf$HeartDisease) = levels(trainMldf$HeartDisease)
yPredRf_proba_old = predict(rf_mod, newdata = testMldf, type='prob')

yPredRf_proba = yPredRf_proba_old[,1]
yPredRf = rep(1,length(yPredRf_proba))
yPredRf[yPredRf_proba > 0.5] = 0
length(yPredRf_proba)

# err_table_rf = table(testMldf$HeartDisease, yPredRf)
# print(err_table_rf)
# test_err_rf = err_table_rf[2,2]/(err_table_rf[1,2]+err_table_rf[2,2])
yPredRf = factor(yPredRf, levels=c(0,1), labels=c("X0","X1"))
mean(yPredRf!=testMldf$HeartDisease)

```


```{r}
cm_rf = confusionMatrix(yPredRf,testMldf$HeartDisease,positive='X1' )
eval_rf = met(cm_rf)
print(cm_rf)
```

```{r}
library(pROC)
roc_rf = roc(testMldf$HeartDisease , yPredRf_proba_old[,2])
plot(roc_rf, main = "ROC Curve for Random Forest")
```


## Decision Tree
```{r}
library(tree)
r_tree = tree(HeartDisease~., data =trainMldf)
summary(r_tree)
```

```{r}
plot(r_tree)
text(r_tree, pretty=3)

```



```{r}
yPredTree_proba_old  = predict(r_tree, newdata = testMldf)
yPredTree_proba  = yPredTree_proba_old
yPredTree_proba = yPredTree_proba[,1]
yPredTree = rep(1,length(yPredTree_proba))
yPredTree[yPredTree_proba > 0.5] = 0
# length(yPredTree_proba)
# 
# err_table_tree = table(testMldf$HeartDisease, yPredTree)
# print(err_table_tree)
# test_err_tree = err_table_tree[2,2]/(err_table_tree[1,2]+err_table_tree[2,2])
# print(test_err_tree)
mean(yPredTree!=testMldf$HeartDisease)
yPredTree = factor(yPredTree, levels=c(0,1), labels=c("X0","X1"))
```


```{r}
cm_dt = confusionMatrix(yPredTree,testMldf$HeartDisease )
eval_dt = met(cm_dt)
print(cm_dt)
```



```{r}

library(pROC)
roc_tree = roc(testMldf$HeartDisease , yPredTree_proba_old[,2])
plot(roc_tree, main = "ROC Curve for Decision Tree")
```


## KNN
```{r}
library(caret)

Xtrain = unalteredDf[,-c(1)]
Xtest = unalteredDf[,-c(1)]
#unalteredDf$HeartDisease = as.factor(unalteredDf$HeartDisease)
ytrain = unalteredDf[,1]
ytest = unalteredDf[,1]

```


```{r}
library(caret)

library(TeachingDemos)
set.seed(char2seed("Priya"))
x = trainControl(method = "cv", number = 3, classProbs = TRUE, summaryFunction = twoClassSummary)

knn_mod <- train(HeartDisease~. , data = trainMldf, method = "knn", preProcess = c("center","scale"), trControl = x,  metric = "ROC")

print(knn_mod)
```




```{r}
plot(knn_mod)
```




```{r}
yPredKNN_proba_old <- predict(knn_mod,testMldf, type = "prob")
yPredKNN_proba <- yPredKNN_proba_old

yPredKNN_proba = yPredKNN_proba[,1]
yPredKNN = rep(1,length(yPredKNN_proba))
yPredKNN[yPredKNN_proba > 0.5] = 0
# length(yPredKNN_proba)
# 
# err_table_KNN = table(testMldf$HeartDisease, yPredKNN)
# print(err_table_KNN)
# test_err_knn = err_table_KNN[2,2]/(err_table_KNN[1,2]+err_table_KNN[2,2])
# print(test_err_knn)
mean(yPredKNN!=testMldf$HeartDisease)
yPredKNN = factor(yPredKNN, levels=c(0,1), labels=c("X0","X1"))
```


```{r}
cm_knn = confusionMatrix(yPredKNN,testMldf$HeartDisease, positive='X1' )
eval_knn = met(cm_knn)
print(cm_knn)
```


```{r}

library(pROC)
roc_knn = roc(testMldf$HeartDisease , yPredKNN_proba_old[,2])
plot(roc_knn, main = "ROC Curve for KNN")
```

## SVM

```{r}
library(e1071)

svm_mod = svm(HeartDisease ~ .,data = trainMldf,kernel = 'linear', probability=TRUE)
summary(svm_mod)

```



```{r}
yPredSVM = predict(svm_mod, newdata = testMldf, probability=TRUE)
yPredSVm_proba_old  = attr(yPredSVM, "probabilities")[,2]
# err_table_SVM = table(testMldf$HeartDisease, yPredSVM)
# print(err_table_SVM)
# 
# test_err_svm = err_table_SVM[2,2]/(err_table_SVM[1,2]+err_table_SVM[2,2])
# print(test_err_svm)
mean(yPredSVM!=testMldf$HeartDisease)



```


```{r}
cm_svm = confusionMatrix(yPredSVM,testMldf$HeartDisease, positive = "X1" )
eval_svm = met(cm_svm)
print(cm_svm)
```

```{r}
library(pROC)
SVMROC <- roc(testMldf$HeartDisease,yPredSVm_proba_old)
plot(SVMROC,  main = "ROC Curve for SVM")
```


```{r} 
eval_acc = c(eval_glm$Acc, eval_rf$Acc,  eval_dt$Acc, eval_knn$Acc, eval_svm$Acc)
models = c("GLM","RF","DT","KNN","SVM")
plot_df = data.frame(
name=models,
value = eval_acc
)
ggplot(plot_df, aes(x=name, y=value, fill=name)) + 
  geom_bar(stat = "identity")+
  labs(title = "Accuracy", y = "Accuracy", fill = "Models")
```
```{r}
eval_sens = c(eval_glm$Sens, eval_rf$Sens,  eval_dt$Sens, eval_knn$Sens, eval_svm$Sens)
eval_spec = c(eval_glm$Spec, eval_rf$Spec,  eval_dt$Spec, eval_knn$Spec, eval_svm$Spec)

plot_eval_df =  data.frame(
Models = models,
Accuracy=eval_acc,
Sensitivity = eval_sens,
Specificity = eval_spec
)
# ggplot(plot_eval_df, aes(fill=models, y=value, x=models)) + 
#     geom_bar(position="dodge", stat="identity")


```


```{r}
plot_eval_data <- data.frame(Value = c(eval_acc, eval_sens, eval_spec),  # Create example data
                   Model = rep(c("GLM",
                                 "RF",
                                 "DT",
                                 "KNN",
                                 "SVM"
                                 ),
                               times = 3),
                   Metric = c(rep(c("Accuracy"),5),
                                rep(c("Sensitivity"),5),
                                rep(c("Specificity"),5)
                                )
)




print(plot_eval_data)

```


```{r}
ggplot(plot_eval_data, aes(fill=Metric, y=Value, x=Model)) + 
    geom_bar(position="dodge", stat="identity")+labs(title = "Evaluation Metrics")
      
```




